Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees

### 2024-02-03 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Paper revisions for BRM
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3279 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ea0ca6a39dd2e0dafa4274567df49d2023c9ee7d

\title{Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity. We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with differently shaped trajectories in linear growth curve models. Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. In simulated and real-world data, we assess the performance of the extensions and compare against other partitioning methods for growth curve models. Extended GLMM trees perform more accurately than the original algorithm and LongCART, and similarly accurate as structural equation model (SEM) trees. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and are much faster to compute.\\}

@@

<<echo=FALSE>>=
library("glmertree")
load("bag_ids, N = 250, Science.Rda")
load("Science ability data.Rdata")
names(sciedata) <- c("CHILDID", "GENDER", "RACE", "SES", "GMOTOR", "FMOTOR", 
                 "INTERN", "EXTERN", "INTERP", "CONTRO", "FIRKDG", 
                 "AGEBASELINE", "science", "score", "asmtmm", "months")
data <- sciedata[sciedata$CHILDID %in% bag_ids[ , 39], ]
data$CHILDID <- factor(data$CHILDID)
data$GENDER <- factor(data$GENDER)
data$asmtmm <- factor(data$asmtmm)
library("lmerTest")
LMM <- lmer(score ~ months + (1|CHILDID), data = data)
icc <- data.frame(VarCorr(LMM))
@

Development over time is of prime interest in psychological research. For example, in educational studies researchers may want to model student's academic development over time; in clinical studies researchers may want to model patients' symptoms over time. Mixed-effects or latent-variable models can be used to model such trajectories and allow for explaining heterogeneity with covariates of a-priori known relevance \citep[e.g., ][]{NeisyMatt18}. However, when these covariates are not known in advance, methods for identifying them are needed.

As an example, trajectories of science knowledge and skills from a sample of 250 children are depicted in Figure~\ref{fig:global_trajectories}. The children were assessed at three timepoints across grades 3 through 8\footnote{Further details on the source of these data are provided in Study~III.}. The red line depicts the estimated average trajectory, while the gray lines depict individual trajectories. The gray lines reveal substantial variability between the children, both in initial levels and growth over time. An obvious research aim would be to identify covariates that can explain or predict this heterogeneity.

\begin{figure}[b]%
\caption{Growth curves of science ability.}
\begin{subfigure}{0.65\textwidth}%
<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
lines(x = c(min(data$months), max(data$months)), 
     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
     lwd = 3, col = 2)
@
\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories, red line depicts average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories.}
\end{subfigure}%
\label{fig:global_trajectories}
\end{figure}%

@@

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope. Groups of children with higher SES also have higher intercepts, indicating higher average science ability. The group of children with lower SES (node~2) is further split based on gross motor skills, with higher motor skills resulting in a higher intercept. The group of children with intermediate levels of SES (node~6) is further split based on internalizing problems, with lower internalizing problems resulting in a higher intercept. The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher ability also gain more ability over time.

\begin{figure}%
\caption{Partitioned growth curves of science ability.}
\begin{subfigure}{0.9\textwidth}%
<<fig=TRUE, height=9, width=13.5, echo=FALSE>>=
LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
testdat$CHILDID <- factor(testdat$CHILDID)
testdat$GENDER <- factor(testdat$GENDER)
testdat$asmtmm <- factor(testdat$asmtmm)
## Remove obs with levels of race in test that are not in train
levs <- unique(testdat$RACE) %in% unique(data$RACE)
if (!all(levs)) {
  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
}

## Fit models
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
icc_r <- as.data.frame(VarCorr(lmm_tree_r))
lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
icc_c <- as.data.frame(VarCorr(lmm_tree_c))
lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
lmm_preds <- predict(LMM, newdata = testdat, re.form=NA)

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\end{subfigure}%
~
\begin{subfigure}{0.25\textwidth}%
\vspace{-2.95cm}
\hspace{-2.5cm}
\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
library("kableExtra")
coefs <- coef(lmm_tree_r)
coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
colnames(coefs) <- c("Node", "Intercept", "Slope")
coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
                                                       coefs$Intercept[!grepl("-", coefs$Intercept)])
kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
        row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
\label{subfig:coefs}
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories. The table presents numerical estimates of fixed intercepts and slopes.}
\label{fig:LMM_tree_r}
\end{figure}%

<<echo=FALSE, results=tex, fig.pos='b'>>=
# library("kableExtra")
# coefs <- coef(lmm_tree_r)
# coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
# colnames(coefs) <- c("Node", "Intercept", "Slope")
# coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
# coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
# coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
#                                                        coefs$Intercept[!grepl("-", coefs$Intercept)])
# kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs",
#                              align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
#         caption = "Fixed-effects coefficient estimates from the terminal nodes of Figure 2.",
#         row.names = FALSE, digits = 3L),
#   font_size = 11, full_width=FALSE)
@

@@

Other methods for partitioning LGCMs take a fully local estimation approach: Within every node or subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that subgroup only. This fully local estimation approach provides more flexibility, but yields higher computational burden and model complexity. In contrast, GLMM trees estimate a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the results. Furthermore, the fully local estimation requires possible partitioning variables to be measured at the highest level of nesting, while GLMM trees' local-global estimation approach allows partitioning variables to be measured at any level.

@@

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
Throughout this paper we focus on the case with a continuous, normally-distributed response $y_i$ with constant variance $\sigma_\epsilon$. Therefore, the identity function can be used for the link $g$ but generalizations to other response variable types within the GLM are straightforward. In the general notation above, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject~$i$ ($i = {1, \dots, N}$), comprising $p$ regressors plus one column of 1s for the intercept. In the following, we assume that time is the predictor variable of interest (i.e., $p = 1$), where the number $n_i$ and spacing of observed timepoints may differ between subjects. The fixed-effects parameters $\beta_j$ (here, intercept and time slope) in GLMM trees are node-specific, i.e., their value depends on the subgroup/node~$j$ into which subject~$i$ falls. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is the corresponding vector of random effects for subject $i$. Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.

@@

\begin{figure}%
\caption{GLMM tree estimated by initializing with zero random effects.}
\begin{subfigure}{1.2\textwidth}
<<fig=TRUE, height=9, width=16, echo=FALSE>>=
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
icc <- as.data.frame(VarCorr(lmm_tree))
lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
@
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree}
\end{figure}%

To illustrate, we applied both initialization approaches to the dataset from Figure~\ref{fig:global_trajectories}. In fact, the tree presented in Figure~\ref{fig:LMM_tree_r} was estimated by initializing with the random effects. Initializing estimation assuming zero random effects resulted in the tree in Figure~\ref{fig:LMM_tree}. The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in  Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills. Considering the relatively large number of subgroups in  Figure~\ref{fig:LMM_tree_r}, some with relatively small sample sizes, this tree may overfit and not generalize well to other samples.

@@

The tree in Figure~\ref{fig:LMM_tree_c} was estimated using cluster-level parameter stability tests. Application of GLMM trees with both cluster-level parameter stability tests and random-effects initialization yielded the exact same tree structure and parameter estimates. Compared to Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, the cluster-level parameter stability tests provided the most parsimonious tree structure thus far. The estimated variances of the random intercept were \Sexpr{formatC(data.frame(VarCorr(lmm_tree_r$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree_r}, \Sexpr{formatC(data.frame(VarCorr(lmm_tree$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree}) and \Sexpr{formatC(data.frame(VarCorr(lmm_tree_c$lmer))$vcov[1], digits = 2, format = "f")} (Figure~\ref{fig:LMM_tree_c}, indicating that the more parsimonious the tree structure, the more variance will be captured by the random effects.

In the next sections, we assess performance of the original algorithm and proposed adaptations through more extensive empirical evaluations.

@@

\FloatBarrier
\section{Study~I: Assessment of Subgroup Recovery}

>>>>

\title{Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}
\authornote{
\textbf{Declarations}

Funding: Not applicable. Conflicts of interest: Not applicable. Ethics approval: Not applicable, simulated and publicly available data from the ECLS-K study were analysed. Consent to participate: Not applicable. Consent for publication: Not applicable. Availability of data and materials: All methods used are implemented in R and associated packages, as described in the Method section. Code availability: Data and R code are available as supplementary materials. None of the experiments were preregistered.}

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity. We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with different trajectories in linear growth curve models. Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. In simulated and real-world data, we assess the performance of the extensions and compare against other partitioning methods for growth curve models. Extended GLMM trees perform more accurately than the original algorithm and LongCART, and similarly accurate as structural equation model (SEM) trees. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and are much faster to compute.\\}

@@

<<echo=FALSE>>=
library("glmertree")
load("bag_ids, N = 250, Science.Rda")
load("Science ability data.Rdata")
names(sciedata) <- c("CHILDID", "GENDER", "RACE", "SES", "GMOTOR", "FMOTOR", 
                 "INTERN", "EXTERN", "INTERP", "CONTRO", "FIRKDG", 
                 "AGEBASELINE", "science", "score", "asmtmm", "months")
data <- sciedata[sciedata$CHILDID %in% bag_ids[ , 39], ]
data$CHILDID <- factor(data$CHILDID)
data$GENDER <- factor(data$GENDER)
data$asmtmm <- factor(data$asmtmm)
library("lmerTest")
LMM <- lmer(score ~ months + (1|CHILDID), data = data)
@

Development over time is of prime interest in psychological research. For example, in educational studies researchers may want to model student's academic development over time; in clinical studies researchers may want to model patients' symptoms over time. Mixed-effects or latent-variable models can be used to model such trajectories and allow for explaining heterogeneity with covariates of a-priori known relevance \citep[e.g., ][]{NeisyMatt18}. However, when these covariates or their shape of association with the response are not known in advance, methods for identifying them are needed.

As an example, trajectories of science knowledge and skills from a sample of 250 children are depicted in Figure~\ref{fig:global_trajectories}. The children were assessed at three timepoints across grades 3 through 8\footnote{Further details on the source of these data and the scaling of time are provided in Study~III.}. The red line depicts the estimated average trajectory, while the gray lines depict individual trajectories. The gray lines reveal substantial variability between the children, both in initial levels and growth over time. An obvious research aim would be to identify covariates that can explain or predict this heterogeneity.

\begin{figure}[b]%
\caption{Growth curves of science ability.}
\begin{subfigure}{0.65\textwidth}%
<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
lines(x = c(min(data$months), max(data$months)), 
     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
     lwd = 3, col = 2)
@
\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories, red line depicts average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here. To facilitate interpretation, the labels on the $x$-axes represent values in the original time metric.}
\end{subfigure}%
\label{fig:global_trajectories}
\end{figure}%

@@

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope, which are presented in Table~\ref{tab:local_coefs}. Groups of children with higher SES also have higher intercepts, indicating higher average science ability. The group of children with lower SES (node~2) is further split based on gross motor skills, with higher motor skills resulting in a higher intercept. The group of children with intermediate levels of SES (node~6) is further split based on internalizing problems, with lower internalizing problems resulting in a higher intercept. The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher baseline ability also gain more ability over time.

\begin{figure}%
\caption{Partitioned growth curves of science ability.}
\begin{subfigure}{1\textwidth}%
<<fig=TRUE, height=9, width=13.5, echo=FALSE>>=
LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
testdat$CHILDID <- factor(testdat$CHILDID)
testdat$GENDER <- factor(testdat$GENDER)
testdat$asmtmm <- factor(testdat$asmtmm)
## Remove obs with levels of race in test that are not in train
levs <- unique(testdat$RACE) %in% unique(data$RACE)
if (!all(levs)) {
  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
}

## Fit models
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\end{subfigure}%
%~
%\begin{subfigure}{0.25\textwidth}%
%\vspace{-2.95cm}
%\hspace{-2.5cm}
%\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
# library("kableExtra")
# coefs <- coef(lmm_tree_r)
# coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
# colnames(coefs) <- c("Node", "Intercept", "Slope")
# coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
# coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
# coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
#                                                        coefs$Intercept[!grepl("-", coefs$Intercept)])
# kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
#         row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
%\label{subfig:coefs}
%\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{months after baseline})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here. To facilitate interpretation, the labels on the $x$-axes represent values in the original time metric.}
\label{fig:LMM_tree_r}
\end{figure}%

<<echo=FALSE, results=tex, fig.pos='b'>>=
library("kableExtra")
coefs <- coef(lmm_tree_r)
coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
colnames(coefs) <- c("Node", "Intercept", "Slope")
coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
                                                      coefs$Intercept[!grepl("-", coefs$Intercept)])
kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs",
                    align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
                    caption = "Estimated fixed-effects coefficients for the terminal nodes of Figure 2.",
                    row.names = FALSE, digits = 3L),
              font_size = 11, full_width=FALSE, position = "left")
@

@@

Other methods for partitioning LGCMs take a fully local estimation approach: Within every node or subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that subgroup only. This fully local estimation approach provides more flexibility, but also higher computational burden and model complexity. In contrast, GLMM trees estimate a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the results. Furthermore, the fully local estimation requires possible partitioning variables to be measured at the highest level of nesting, while GLMM trees' local-global estimation approach allows partitioning variables to be measured at any level.

@@

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%

Throughout this paper we focus on the case with a continuous, normally-distributed response $y_i$ with constant variance $\sigma_\epsilon^2$. Therefore, the identity function can be used for the link $g$ but generalizations to other response variable types within the GLM are straightforward. In the general notation above, $X_i$ is the $n_i \times (p + 1)$ fixed-effects design matrix for subject i $(i = 1, \dots, N)$, comprising $p$ regressors plus one column of 1s for the intercept. In this paper, we assume that time is the only time-varying covariate of interest (i.e., $p = 1$), but other time-varying covariates can also be included in $X_i$. The number $n_i$ and spacing of observed timepoints may differ between subjects. Ttime-constant covariates are not contained in $X_i$, but enter the model through the subgroup indicator $j$. The value of the fixed-effects parameters $\beta$ (here, intercept and time slope) in GLMM trees are node-specific, their value depending on the subgroup/node $j$ into which subject $i$ falls. Subgroup membership is defined by the values of time-constant covariates\footnote{In principle, it is also possible to use time-varying covariates for partitioning with GLMM trees, but this might not be appropriate for partitioning growth curves, as participants changing subgroups over time yields discontinuous growth curves.}.

As in the standard GLMM \citep[e.g., ][]{PinhyBate00}, GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of Xi, and bi is the corresponding vector of random effects for subject $i$. Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.

@@

\begin{figure}%
\caption{GLMM tree estimated by initializing with zero random effects.}
\begin{subfigure}{1.2\textwidth}
<<fig=TRUE, height=9, width=16, echo=FALSE>>=
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
vcov_d <- as.data.frame(VarCorr(lmm_tree))
@
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree}
\end{figure}%

To illustrate, we applied both initialization approaches to the dataset from Figure~\ref{fig:global_trajectories}. In fact, the tree presented in Figure~\ref{fig:LMM_tree_r} was estimated by initializing with the random effects. Initializing estimation assuming zero random effects resulted in the tree in Figure~\ref{fig:LMM_tree}. The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in  Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills. Considering the relatively large number of subgroups in  Figure~\ref{fig:LMM_tree_r}, some with relatively small sample sizes, this tree may overfit and not generalize well to other samples\footnote{Generalizability will be empirically evaluated using cross validation in Study~III.}.

@@

The tree in Figure~\ref{fig:LMM_tree_c} was estimated using cluster-level parameter stability tests. Combining cluster-level parameter stability tests and random-effects initialization yielded the exact same tree structure and parameter estimates. Compared to the trees in Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, the cluster-level parameter stability tests provided the most parsimonious tree structure thus far. Table~\ref{tab:ranef} presents the estimated random-effects parameters for the models fitted thus far. Not surprisingly, the estimated residual standard deviation ($\hat{\sigma}_\epsilon$) differs only very slightly between models. Yet, substantial differences can be seen for the standard deviation of the random intercept ($\hat{\sigma}_\epsilon$): With more subgroups or terminal nodes, less variance is captured by the random effects and more variance is captured by the tree structure.

<<echo=FALSE>>=
vcov_d <- as.data.frame(VarCorr(lmm_tree))
vcov_r <- as.data.frame(VarCorr(lmm_tree_r))
vcov_c <- as.data.frame(VarCorr(lmm_tree_c))
vcv <- as.data.frame(VarCorr(LMM))
tab <- data.frame(Figure = 1:4,
                  Model = c("LMM", rep("LMM tree", times = 3)), 
                  Estimation = c("default", "random effects initialization", 
                                 "default", "clustered covariances"),
                  n_subs = c(1L, (length(lmm_tree_r$tree)+1)/2, 
                             (length(lmm_tree$tree)+1)/2, (length(lmm_tree_c$tree)+1)/2))
tab$sig_b <- c(vcv[vcv$grp == "CHILDID", "sdcor"], vcov_r[vcov_r$grp == "CHILDID", "sdcor"], 
               vcov_d[vcov_d$grp == "CHILDID", "sdcor"], vcov_c[vcov_c$grp == "CHILDID", "sdcor"])
tab$sig_epsb <- c(vcv[vcv$grp == "Residual", "sdcor"], vcov_r[vcov_r$grp == "Residual", "sdcor"], 
               vcov_d[vcov_d$grp == "Residual", "sdcor"], vcov_c[vcov_c$grp == "Residual", "sdcor"])
tab$pool_sd <- c(sqrt(sum(vcv$vcov)), sqrt(sum(vcov_r$vcov)),
                 sqrt(sum(vcov_d$vcov)), sqrt(sum(vcov_c$vcov)))
ds <- dist(coef(lmm_tree_r)[ , "(Intercept)"], method = "manhattan") / tab$pool_sd[tab$Figure == 1L]
@

The variances in Table~\ref{tab:ranef} can also be used to quantify variability associated with the trajectories in the terminal nodes of the trees. The third column of Table~\ref{tab:ranef} presents pooled standard deviations that can be computed from variance estimates \citep{WestyKenn14}. These allow, for example, to gauge the magnitude of subgroup differences in the subgroup-specific intercepts in Table~\ref{tab:local_coefs}, which range from $d = \Sexpr{round(min(ds), digits = 2)}$ to $d = \Sexpr{round(max(ds), digits = 2)}$. Yet, these estimates are based on a modest sample size, and are based on training data only, so can be overly optimistic. In the next sections, we evaluate performance on simulated (Studies~I~and~II) and real data (Study~III).

<<echo=FALSE, results=tex, fig.pos='b'>>=
vcov_d <- as.data.frame(VarCorr(lmm_tree))
vcov_r <- as.data.frame(VarCorr(lmm_tree_r))
vcov_c <- as.data.frame(VarCorr(lmm_tree_c))
vcv <- as.data.frame(VarCorr(LMM))
tab <- data.frame(Figure = 1:4,
                  Model = c("LMM", rep("LMM tree", times = 3)), 
                  Estimation = c("default", "random effects initialization", 
                                 "default", "clustered covariances"),
                  n_subs = c(1L, (length(lmm_tree_r$tree)+1)/2, 
                             (length(lmm_tree$tree)+1)/2, (length(lmm_tree_c$tree)+1)/2))
tab$sig_b <- c(vcv[vcv$grp == "CHILDID", "sdcor"], vcov_r[vcov_r$grp == "CHILDID", "sdcor"], 
               vcov_d[vcov_d$grp == "CHILDID", "sdcor"], vcov_c[vcov_c$grp == "CHILDID", "sdcor"])
tab$sig_epsb <- c(vcv[vcv$grp == "Residual", "sdcor"], vcov_r[vcov_r$grp == "Residual", "sdcor"], 
               vcov_d[vcov_d$grp == "Residual", "sdcor"], vcov_c[vcov_c$grp == "Residual", "sdcor"])
tab$pool_sd <- c(sqrt(sum(vcv$vcov)), sqrt(sum(vcov_r$vcov)),
                 sqrt(sum(vcov_d$vcov)), sqrt(sum(vcov_c$vcov)))
colnames(tab)[4:7] <- c("# of nodes", "$\\hat{\\sigma}_b$", "$\\hat{\\sigma}_\\epsilon$", "$\\hat{\\sigma}_{\\text{pooled}}$")
kable_styling(
  kable(tab, format = "latex", linesep = "", row.names = FALSE, 
        label = "ranef", digits = 3, align = c("cllccc"), booktabs = TRUE, escape =FALSE, 
        caption = "Estimated random-effects parameters for fitted mixed-effects models in Figures 1 through 4."),
  position = "left")
@

<<echo=FALSE, eval=FALSE>>=
test_dat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
test_dat$CHILDID <- factor(test_dat$CHILDID)
test_dat$GENDER <- factor(test_dat$GENDER)
test_dat <- test_dat[test_dat$RACE != 6, ]
lmmt_preds <- predict(lmm_tree, re.form = NA, newdata = testdat)
lmmt_c_preds <- predict(lmm_tree_c, re.form = NA, newdata = testdat)
lmmt_r_preds <- predict(lmm_tree_r, re.form = NA, newdata = testdat)
mean((test_dat$score - lmmt_preds)^2)
mean((test_dat$score - lmmt_r_preds)^2)
mean((test_dat$score - lmmt_c_preds)^2)
var(test_dat$score)
lmm_preds <- predict(LMM, newdata = test_dat, re.form = NA)
mean((test_dat$score - lmm_preds)^2)
@

@@

\FloatBarrier
\section{Study~I: Assessment of Subgroup Recovery}
\label{sec:studyI}
